Complexity-stability relationships disordered dynamical systems

### 2024-06-24 papers/response-referee.md

We thank the referee for the thoughtful review of our work and for raising important points, addressed below, which we think helped us improve the paper.

- *The claim is probably true, but unfortunately it does not describe a generic situation, because it is very hard to imagine that the exponents alpha, beta, gamma used in the model are identical for all populations. Perhaps the model would be more interesting if it included a spectrum of different exponents, described, for example, by a probabilistic law.*

We agree that indeed a spectrum of different exponents is probably more realistic, and thank the reviewer for pointing that out.

We consider now the case in which the exponents are not identical for every variable $x_i$ but instead we have them from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.

We now added a Supplemental Material document where report results of simulations showing that our results are robust as long as $\sigma_e$ is small enough that $\displaystyle{\min_i\beta_i>\max_i\alpha_i}$. We also added a paragraph in the main text where we discuss this point.

@@

- *The way of writing the matrix elements in Equations (7) is somewhat confusing. For clarity, the Kronecker delta can be used on the right side of these equations.*

We modified Equations (7) accordingly.

- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>. The self-consistency condition requires that the probability distribution (12) has exactly the same expected values. This should be discussed in detail.*

We agree with the referee that the discussion about the expected values and self-consistency conditions needed clarifications.

In some cases, for example, $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations. However in potential cases of interest, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ characterized by a distribution with a power-law tail with exponent $-3/2$, the moments diverge. We now modified the discussion removing the Gaussian approximation and considering the introduction of a cut-off. The idea behind this cut-off is described in the main text and in the Supplemental Material.

@@

## Referee B

We thank the referee for their thorough and critical review of our work. We address each of the comments below point by point. We believe the results of our paper are more transparent thanks to these modifications.

@@

- *Modelling Framework Justification:*

     *The rationale behind the proposed generalization of the GLV model is claimed to be "natural" based on physical and biological grounds, specifically using power laws. Nevertheless, the literature extensively demonstrates species abundance growth being well-described by logistic equations or oscillatory predator-prey equations, that indeed are linear (except for the carrying capacity term). The justifications provided are weak, with the provided examples failing to convincingly represent the complex dynamics of natural systems, such as forest biomass distribution. If we think to a forest there are various heights of trees, and in general the biomass in not uniformly distributed, and thus also the growth is not necessary constrained by the surface. The sublinear scaling of the “production function”, f, if this is really the right conclusion based on the references of the manuscript, must be the result of a cooperative behavior of the interacting system and as such it should not appear as hand-added like the authors do. Furthermore, in reference 27 the sublinear scaling with k=3/4 refers to the scaling of the single individual growth rather than to a community.*

We recently published a separate paper (cited as Ref. [17]) to outline in some detail the ecological motivations and consequences of competitive models with sublinear production. The goal of the present letter is different. Following Refs. [18] and [19], we show in a minimal setting (and for an audience of physicists) the existence of the two opposite complexity-stability relationships in disordered dynamical systems. We have reword

We are aware that reference [27] refers to the scaling of single individuals. We now mention that explicitly.

@@

- *Lack of Proper Scaling:*

    *There is no proper scaling of the interactions, so the thermodynamic limit is not well defined. And this is not discussed or compared with other models in the literature applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv preprint arXiv:2306.11031, 2023). Relatedly, the comparison with GLV with quenched disorder (e.g. “The properties of large random dynamical systems is often portrayed in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the interactions strengths with the system size.*

This is a delicate point and we thank the referee for raising it. This gives us the opportunity to elaborate more on the choice of not scaling the interactions.

As also discussed in reference [25] (end of "I. MODEL DEFINITION"), the choice of scaling or not the interaction depends on the ecological setting of interest. 
We believe that it is appropriate to fix the strength and heterogeneity of the interaction ($mean(A)=\mu$ and $var(A)=\sigma^2$) if our goal is to study the effects of changes in the number of degrees of freedom.
In this setting, while the full limit in which $N\to\infty$ leads to the homogeneous interaction case with $x^*\to0$ due to competitive pressure, the cavity approximation does a good job at describing the empirical distributions obtained from simulations as long as $\sigma\sqrt{N}$ is not negligible with respect to $\mu N$.

### 2024-06-25 papers/response-referee.md

- *The way of writing the matrix elements in Equations (7) is somewhat confusing. For clarity, the Kronecker delta can be used on the right side of these equations.*

We thank the referee for this comment, we modified Equations (7) accordingly.

- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>. The self-consistency condition requires that the probability distribution (12) has exactly the same expected values. This should be discussed in detail.*

We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarification.

@@

We thank the referee for the thorough and critical review. We address each of the comments below point by point. We believe the results of our paper are now more transparent thanks to the work needed to address the points raised by the referee.

- *Modelling Framework Justification:*

     *The rationale behind the proposed generalization of the GLV model is claimed to be "natural" based on physical and biological grounds, specifically using power laws. Nevertheless, the literature extensively demonstrates species abundance growth being well-described by logistic equations or oscillatory predator-prey equations, that indeed are linear (except for the carrying capacity term). The justifications provided are weak, with the provided examples failing to convincingly represent the complex dynamics of natural systems, such as forest biomass distribution. If we think to a forest there are various heights of trees, and in general the biomass in not uniformly distributed, and thus also the growth is not necessary constrained by the surface. The sublinear scaling of the “production function”, f, if this is really the right conclusion based on the references of the manuscript, must be the result of a cooperative behavior of the interacting system and as such it should not appear as hand-added like the authors do. Furthermore, in reference 27 the sublinear scaling with k=3/4 refers to the scaling of the single individual growth rather than to a community.*

We understand the criticism of the referee, pointing out the need for better justification of models of this kind in ecology. Our goal in this letter, however, is not to propose a model for a specific ecological scenario. Rather, we employ this model for complex systems (following references [18] and [19]) to show in a minimal setting the existence of the two opposite complexity-stability relationships in disordered dynamical systems, possibly with applications beyond ecology.

Nevertheless, the examples we provide are indicative that models of this class can be relevant in ecology (see reference [17], where a discussion of possible mechanisms is present; even though admittedly more work is required in that direction, which is beyond the scope of our paper).

### 2024-06-25 papers/submission/beyond-may.tex
update draft and response

However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,mazzarisi2024}.

@@

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case the system is always feasible and the distribution can be normalized. However, the first and second moments formally diverge.
This is problem can be overcome considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$. In orther words such that there is less 1 variable with density above $\Lambda$.

@@

Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$. (Numerically, we define $\sigma_c$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model. Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

>>>>

However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, at the individual organism level (populations of cells), growth has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities~\cite{Hatton2015,Hatton2024}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law, i.e. $g(x) \sim h(x) \sim x^{1/2}$~\cite{Barbier2021,mazzarisi2024}.

@@

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case the system is always feasible and the distribution can be normalized. However, the first and second moments formally diverge.
This is problem can be overcome considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$. In orther words such that there is less 1 variable with density above $\Lambda$ (see Supplemental Material).

@@

Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$. (Numerically, we define $\sigma_c$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model. Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

Finally, we check the robustness of our findings in the scenario in which the exponents are not the same for every variable $x_i$. As a test, we can extract them from different Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$. 
Simulations show (see Supplemental Material) that our results hold as long as $\sigma_e$ is small enough, in particular, we observe a loss of stability when $\min_i\beta_i>\max_i\alpha_i$.
